---
title: "Speed_Precip_Compliance"
author: "Joshua Ferreri"
date: "April 10, 2016"
output: html_document
---

#Compliance of Wind Speed conversions:

##Knots to other speed units:

Testing compliance of Wind Speed conversions for knots to other units of measure using the `knots_to_speed` function in the `weathermetrics` package with the National Oceanic and Atmospheric Administration's [online wind speed converter](http://www.srh.noaa.gov/epz/?n=wxcalc_windconvert) (accessed February 22, 2016).

The following examples convert from 10 knots to mph, mps, fps, and kmph. 

```{r}
#knots to mph
mph <- knots_to_speed(10, "mph", round.out = TRUE)
mph
```

The value from the above example using the `knots_to_speed` function, `r mph` mph, complies with that obtained using the NOAA converter, which returns a value of 11.5 mph for a value of 10 knots.

```{r}
#knots to mps
mps <- knots_to_speed(10, "mps", round.out = TRUE)
mps
```

The value from the above example using the `knots_to_speed` function, `r mps`mps, complies with that obtained using the NOAA converter, which returns a value of 5.1 mps for a value of 10 knots.

```{r}
#knots to ftps
ftps <- knots_to_speed(10, "ftps", round.out = TRUE)
ftps
```

The value from the above example using the `knots_to_speed` function, `r ftps` ftps, complies with that obtained using the NOAA converter, which returns a value of 16.9 ftps for a value of 10 knots.

```{r}
#knots to kmph
kmph <- knots_to_speed(10, "kmph", round.out = TRUE)
kmph
```

The value from the above example using the `knots_to_speed` function, `r kmph` kmph, complies with that obtained using the NOAA converter, which returns a value of 18.5 kmph for a value of 10 knots.

##Common speed units to knots:

Testing compliance of Wind Speed conversions for common wind speed units (mph, mps, ftps, kmph) to knots using the `speed_to_knots` function in the `weathermetrics` package with the National Oceanic and Atmospheric Administration's [online wind speed converter](http://www.srh.noaa.gov/epz/?n=wxcalc_windconvert) (accessed February 22, 2016).

The following examples convert from 10 mph, mps, fps, and kmph, to knots. 

```{r}
#mph to knots
knots <- speed_to_knots(10, "mph", round.out = TRUE)
knots
```

The value from the above example using the `speed_to_knots` function, `r knots` knots, complies with that obtained using the NOAA converter, which returns a value of 8.7 knots for a value of 10 mph.

```{r}
#mps to knots
knots <- speed_to_knots(10, "mps", round.out = TRUE)
knots
```

The value from the above example using the `speed_to_knots` function, `r knots` knots, complies with that obtained using the NOAA converter, which returns a value of 19.4 knots for a value of 10 mps.

```{r}
#ftps to knots
knots <- speed_to_knots(10, "ftps", round.out = TRUE)
knots
```

The value from the above example using the `speed_to_knots` function, `r knots` knots, complies with that obtained using the NOAA converter, which returns a value of 5.9 knots for a value of 10 ftps.

```{r}
#kmph to knots
knots <- speed_to_knots(10, "kmph", round.out = TRUE)
knots
```

The value from the above example using the `speed_to_knots` function, `r knots` knots, complies with that obtained using the NOAA converter, which returns a value of 5.4 knots for a value of 10 kmph.

##Example converting from one speed unit, through knots, to another:

Testing compliance of Wind Speed conversions for common wind speed units (mph, mps, ftps, kmph) to another common wind speed unit converting through knots, using the `speed_to_knots` and `knots_to_speed` functions in the `weathermetrics` package with the National Oceanic and Atmospheric Administration's [online wind speed converter](http://www.srh.noaa.gov/epz/?n=wxcalc_windconvert) (accessed February 22, 2016).

The following examples converts 10 mph to kmph, converting through knots.

```{r}
knots <- speed_to_knots(10, "mph", round.out = FALSE)
kmph <- knots_to_speed(knots, "kmph", round.out = TRUE)
kmph
```

The value from the above example using the `speed_to_knots` and `knots_to_speed` functions to convert from 10 mph to kmph, `r kmph` kmph, complies with that obtained using the NOAA converter, which returns a value of 16.1 kmph for a value of 10 mph.
